Ornaments as indicators of social changes and cultural practices in northeastern Taiwan before and after European colonial period

### 2019-12-10 analysis/paper/paper.Rmd
editted the background section

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages (@Liu2011; @Chen2017). It is generally accepted that Taiwan entered the historical period since the early 17th century due to the colonial presence of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, Taiwan was incorporated into Qing dynasty and a large wave of Han Chinese migrated to Yilan during the late 18th century.

The Spanish founded Fort San Salvador at Keelung in 1626 and Fort San Domingo in 1629 at Tamsui, and sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343]. A report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of indigenous people to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in the Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and the fragmented nature of texts selected by written record keepers that might lead to misinterpretations of artifacts [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources lacking contexts are the problems need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicates multiple origins more likely from Southeast Asia [@Chen2011]. There is no clear evidence shows the European origin of beads, however, a large amount of golden beads at Kiwulan might be introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Based on the archaeological evidence and historical records, Northeastern Taiwan was involved in the regional network with East Asia in the late Iron age that Chinese merchants traded metal materials, clothes, and beads with local indigenous people in Taiwan for local resources and later was incorporated into more international scale due to the arrival of the Europeans. The common finds of foreign stoneware jars from Southeast Asia in local sites were frequently found in European shipwrecks that indicates the direct or indirect interaction. Despite the origin of some ornaments were traced to China, it is likely the large amount of them found at the 17th century sites were traded into due to indirect economic activities stimulated by the Europeans or direct European colonial activities.

@@

Kiwulan (Figure \@ref(fig:KWL-map)) was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The excavation area includes eight open area sections of 262 squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Ornaments were found in different archaeological contexts including post hole area, burials, and middens while digging and some small beads were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest section, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site.

The chronology of Kiwulan can be divided into two phases represented by a upper component (1400-1900 AD) and a lower component (700-1200 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. The relationship between upper component and lower component is still under debate and the sterile might result from dry weather according to a pollen study [@Lin2015; @Chiu2004; @Chen2007]. This paper focuses on the upper component covering a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researches divided the upper component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. To answer our research question, we re-examined and assigned previous layers to a new chronology. For each sampled square we separated excavation units into three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. A large amount of light gray glazed jars were found on the Spanish shipwreck San Diego sinking in 1600 AD, which is believed to be the earliest evidence yet found [@Hsieh2009; @Hsieh1995]. The origin of light gray glazed jars was thought from Southeast China and commonly found at sites in Taiwan that were associated with the European activities, such as Zeelandia fort site in Tainan [@Wang2007]. Regarding large dark brown glazed stoneware jars, they might be made in Southeast Asia and are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. Those jars might be directly imported to Kiwulan by the Europeans to Kiwulan or indirectly by Chinese merchants or other indigenous groups via the regional network in north Taiwan. Either way indicates that the Europeans played an important role in introducing those jars to Taiwan. We used the presence and absence of light grey glazed jar and large stoneware jars, together with the radiocarbon dates, to identify excavation units associated with the pre-European and the start of the European influence at Kiwulan.

The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrant in 1768 [@Chen1963; @Ke1993].

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened postholes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). These postholes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction across the sampling area. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes is 3,173. The use of ornaments as body adornments in burials will be discussed to understand the function of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strands or patterned bands, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports the argument that ornaments were treated as high-value items at Kiwulan based on a few finds in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place across different time periods as it can provide key information concerning social inequality indicated by uneven access to high-value items in relation to the foreign influences.

>>>>

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Liu2011; @Chen2017]. It is generally accepted that Taiwan entered the historical period since the early 17th century due to the colonial presence of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural harbours, northeast Taiwan was involved in a regional trade network through cross culture interactions with Chinese merchants since 14th century, and later the global trade network with the Europeans in the 17th century who brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although on the peripheral location of the trade center, northeastrn Taiwan- Yilan was connected to the trade networks via visits of other indigenous groups, Chineses merchants, and the Europeans via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Keelung in 1626, and Fort San Domingo in 1629 at Tamsui. They sent missionaries to local indigenous settlements in this region for religious purposes [@Blusse2000, pp. 343] who left many records about the life of indigenous. Recording the trade beads, a report by a Dominican priest in 1632 mentioned that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers learned to use carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans (majuorbol) in the tribe would use carnelian beads as magical items in ritual practice for body healing [@Borao2009, pp. 122-151]. Records of an indigenous funeral process document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate a family’s higher prestige [@LiandWu2006, pp. 153]. These historical accounts indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on the demands of indigenous people to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the Dutch feudal system resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan, northeast Taiwan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. In sum, we find multiple sources describing how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer mentions of beads in the Chinese period and the descriptions are limited to clothing, but generally confirm the role of beads as status markers. However, we must recognize the inherent biases in historical records and the fragmented nature of texts generated by the writers [@Galloway2006]. Inconsistent and incomplete descriptions from different written sources need to be compared and examined by studying the ornaments in archaeological contexts.

Currently the ornaments found in northeastern Taiwan in the early historical period, including glass beads, stone beads, and metal ornaments, were considered as imported materials from other regions due to a lack of evidence of beadmaking waste, metalworking, or accessible local raw materials. The chemical composition analysis for glass beads shows high content of lead, coupled with the winding/folding technique, suggest a Chinese beadmaking tradition [@Cheng2008; @Gan2006; @Wang2018]. Although there is a wide variety of metal ornaments such as bells, bracelets, rings, and pendants, the common compositions of metal ornaments are brass and copper with a small number made from lead and tin that indicates multiple origins more likely from Southeast Asia [@Chen2011]. There is no clear evidence showing the European origin of beads, however, a large amount of golden beads at Kiwulan might be introduced by the Spanish through economic activities because similar beads were found at Luzon, northern Philippines, a part of the trading route of the Spanish between 16-19th century [@Wang2007]. Both archaeological evidence and historical records indicating northeastern Taiwan was involved in the regional network with East Asia in the late Iron age that Chinese merchants traded metal materials, clothes, and beads with local indigenous people in Taiwan for local resources and later incorporated into more international scale due to the arrival of the Europeans. The foreign stoneware jars frequently found in European shipwrecks were also commonly found from many sites in Taiwan that suggests a direct or indirect interaction. Despite the origin of some ornaments were traced to China, it is likely that a large amount of ornament found at the 17th century sites were an influence of the Europeans colonial and economic activities.

@@

Kiwulan (Figure \@ref(fig:KWL-map)) was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The excavation includes eight open areas in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artefacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-tern settlement. The center of the site is the open area consists of A and D section which is also the study area where our samples selected from. In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction (Figure \@ref(fig:KWL-sam-area)).

The chronology of Kiwulan can be divided into two phases represented by a upper component (1400-1900 AD) and a lower component (700-1200 AD) separated by a sterile. This chronology is based on the stratigraphy and a series of 32 radiocarbon dates previously published by @Chen2007. The relationship between upper component and lower component is still under debate and the sterile might result from dry weather according to pollen analyses [@Lin2015; @Chiu2004; @Chen2007]. This paper focuses on the upper component covering a sequence spanning pre-European contact, European presence, and Chinese presence. Previous researches divided the upper component into six layers spanning from the 14th century to the 19th century with hundred year intervals according to the radiocarbon dates, excavation depth, consistency of contexts, and types of diagnostic porcelains such as blue and white porcelains [@Hsieh2009; @Wang2011]. To answer our research question, we re-examined and assigned previous layers to a new chronology with three episodes of time: pre-European period, European period, and Chinese period to indicate the dominant foreign influence and cultural interaction for different time periods.

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. A large amount of light gray glazed jars were found on the Spanish shipwreck San Diego sinking in 1600 AD, which is believed to be the earliest evidence yet found [@Hsieh2009; @Hsieh1995]. The origin of light gray glazed jars was thought from Southeast China and commonly found at sites in Taiwan that were associated with the European activities, such as Zeelandia fort site in Tainan [@Wang2007]. Regarding large dark brown glazed stoneware jars, they might be made in Southeast Asia and are frequently found in European shipwrecks from this period for transporting water, wine or other fluids on the long voyages. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. They might be directly imported to Kiwulan by the Europeans to Kiwulan or indirectly by Chinese merchants or other indigenous groups via the regional network in north Taiwan. Either way indicates that the Europeans played an important role in introducing foreign jars to Taiwan. Those jars were widely distributed across the site and served as indicators, together with the radiocarbon dates, to identify the excavation unites associated with the pre-European and the start of the European influence at Kiwulan. The archaeological signature of the Chinese period at Kiwulan is the large amount and diversity of Chinese porcelains in many forms such as bowls, plates, and cups. Other indicators include opium pipe-bowls and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Chinese migrations to Yilan were also recorded in official Chinese records written in the early 19th century recording the first immigrant in 1768 [@Chen1963; @Ke1993].

@@

Ornaments were found in different archaeological contexts including post-holes area, burials, and middens while digging and some small beads were collected through screens with 2 mm and 1.5 mm mesh. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aliened post-holes with in-situ posts (Figure \@ref(fig:KWL-sam-area)). They were sampled because those units were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 30 burials in the sampling area account for one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, and the total number including different ornament classes is 3,173. The use of ornaments as body adornments in burials will be discussed to understand the function of ornaments, but not statistically analysed since the number of ornaments is highly affected by the presence of bead strands or patterned bands, which sometimes contains thousands of beads in an individual burial [@Chen2007]. In addition to ornaments from the dwelling place and burial, there are 27 ornaments found in middens, which to some extent supports the argument that ornaments were treated as high-value items at Kiwulan based on a few finds in the midden contexts compared to high frequency in burial contexts. It is important to analyse the detailed distribution of ornaments from the dwelling place across different time periods as it can provide key information concerning social inequality indicated by uneven access to high-value items in relation to the foreign influences.
